---
title: "dif-in-dif"
output: html_document
date: "2026-06-08"
---

# NOTES
> Need to figure out the best way to handle `id_var`
> Need to add notes around updating STEP 2b
> Need to QA - i get different results with `country` vs `id_numeric`

## SETUP

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(foreign, dplyr, ggplot2, tidyr, scales, did)

# Further colors
ftr_lt_blue <- "#10cee8"
ftr_md_blue <- "#095fbf"
ftr_dk_blue <- "#0f0f2a"
ftr_mint <- "#00f1a1"
ftr_red <- "#f63241"

# Helper function to interpret regression results
interpret_did <- function(didreg, term_name = "treated:time") {
  # 1. Extract the coefficients matrix from the summary object
  coef_matrix <- summary(didreg)$coefficients
  
  # Check if the term exists in the model output
  if (!term_name %in% rownames(coef_matrix)) {
    stop(paste("The term '", term_name, "' was not found in your regression output."))
  }
  
  # Extract point estimate and p-value
  point_estimate <- coef_matrix[term_name, "Estimate"]
  p_value <- coef_matrix[term_name, 4] # The 4th column contains the p-value
  
  # 2. Determine the custom significance level and the corresponding alpha
  if (p_value < 0.01) {
    sig_text <- "1% level (p < 0.01)"
    conf_level <- 0.99
  } else if (p_value < 0.05) {
    sig_text <- "5% level (p < 0.05)"
    conf_level <- 0.95
  } else if (p_value < 0.10) {
    sig_text <- "10% level (p < 0.10)"
    conf_level <- 0.90
  } else if (p_value < 0.15) {
    sig_text <- "15% level (p < 0.15)"
    conf_level <- 0.85
  } else {
    sig_text <- "not statistically significant even at the 15% level (p >= 0.15)"
    conf_level <- 0.85 # Default to an 85% range for reference
  }
  
  # 3. Calculate the confidence interval range based on that significance level
  ci_bounds <- confint(didreg, parm = term_name, level = conf_level)
  ci_lower <- ci_bounds[1]
  ci_upper <- ci_bounds[2]
  
  # 4. Print the plain English interpretation
  cat("==================================================================\n")
  cat("             DIFFERENCE-IN-DIFFERENCES INTERPRETATION            \n")
  cat("==================================================================\n\n")
  
  # Output 1: Significance
  cat("1) Significance of the Impact:\n")
  cat(paste0("   The p-value for the treatment impact is ", round(p_value, 4), ".\n"))
  cat(paste0("   Therefore, the result is statistically significant at the ", sig_text, ".\n\n"))
  
  # Output 2: Point Estimate
  cat("2) Point Estimate of the Impact:\n")
  cat(paste0("   The point estimate is ", format(point_estimate, scientific = FALSE, big.mark = ","), ".\n"))
  cat(paste0("   This means that the treatment caused an average change of \n",
             "   ", format(point_estimate, scientific = FALSE, big.mark = ","), " units in the treatment group per unit of time, \n",
             "   relative to how the control group changed over the same time period.\n\n"))
  
  # Output 3: Range of Impact
  cat("3) Range of Impact:\n")
  cat(paste0("   Based on the significance level determined above, we look at the ", (conf_level * 100), "% confidence interval.\n"))
  cat(paste0("   We can be ", (conf_level * 100), "% confident that the true effect lies in the range between:\n"))
  cat(paste0("   Lower bound: ", format(ci_lower, scientific = FALSE, big.mark = ","), "\n"))
  cat(paste0("   Upper bound: ", format(ci_upper, scientific = FALSE, big.mark = ","), "\n"))
  cat("==================================================================\n")
}
```

## DEFINE VARIABLES

```{r}
t_start <- 1994  # start date of the treatment/intervention
t_var <- "year"  # date column name
d_var <- "y"     # The dependent variable (e.g., revenue, employment)
id_var <- "country" # The unique identifier for units (e.g., customer_id, store_id)
```

## STEP 1: Read in data

```{r}
df <- read.dta("https://www.princeton.edu/~otorres/Panel101.dta")
```

## STEP 2: Create new variables

Once our data is read in, we'll need to create a dummy variable to indicate the time when the treatment started. 

### STEP 2a: Create `post period` dummy variable

For this example, lets assume that treatment started in 1994. In this case, years before 1994 will have a
value of 0 and 1994+ a 1. If you already have this skip this step.

```{r post_dummy, echo=FALSE}
df$time <- ifelse(df[[t_var]] >= t_start, 1, 0)
```

### STEP 2b: Create `treatment group` dummy variable

Next, we need to create a dummy variable to identify the group exposed to the treatment. 

In this example, lets assume:
 > Countries with code 5, 6, and 7 were treated (=1)
 > Countries with code 1-4 were not treated (=0)

```{r treat_dummy, echo=FALSE}
df$treated <- ifelse(
  df$country == "E" |
  df$country == "F" |
  df$country == "G", 1, 0)
```

### STEP 2c: Create `treatment start` variable

Once we have `time` and `treated` created, we can now create the `gname` variable, which will be used in our assessment of the parallel trends assumption. This `gname` variable simply returns the time period for which the treatment started in treated groups, or 0 or non-treated groups.

```{r}
df$gname <- t_start * df$treated
```

## STEP 3: Validate parallel trends assumption

<explain this step>

```{r}
df$id_numeric <- as.numeric(as.factor(df[[id_var]]))

# 1. Run the Group-Time Average Treatment Effects estimation
# Note: 'xformla = ~1' assumes no covariates. Change to '~ var1 + var2' if conditioning on covariates.
did_model <- att_gt(
  yname   = d_var,
  tname   = t_var,
  idname  = "id_numeric",
  gname   = "gname",
  xformla = ~1, 
  data    = df
)

# 2. View the statistical summary (Includes the Parallel Trends Wald Pre-Test P-value)
summary(did_model)

# 3. Plot the group-time treatment effects to visually inspect pre-trends (Red dots)
ggdid(did_model)
```

## STEP 4: Estimating the DID Estimator

With all our new variables defined, it's now time to get the point estimate of the impact and significance of that estimate. The coefficient for ‘did’ is the differences-in-differences estimator.

Note that when evaluating the estimate, we need to focus on the interaction (`treated:time` in this example).

In this example, the effect is significant at 10% with the treatment having a negative effect.

```{r did_estimator, echo=FALSE}
didreg <- lm(y ~ treated * time, data = df)
summary(didreg)
```

## STEP 5: Interpret the results

This script includes a helper function to assist with interpreting the results in a way that can resonate with non-technical audiences. You will likely still need to tweak to suit your needs, but this will provide a solid starting point. 

```{r, echo=FALSE}
# Regression Interpretation
print(interpret_did(didreg))

# Build Comparison Table
means <- df %>%
  group_by(treated, time) %>%
  summarize(mean_y = mean(y, na.rm = TRUE), .groups = 'drop')

## Extract specific mean values for calculations
ctrl_pre  <- means %>% filter(treated == 0, time == 0) %>% pull(mean_y)
ctrl_post <- means %>% filter(treated == 0, time == 1) %>% pull(mean_y)
tr_pre    <- means %>% filter(treated == 1, time == 0) %>% pull(mean_y)
tr_post   <- means %>% filter(treated == 1, time == 1) %>% pull(mean_y)

## Calculate across-time and across-group differences
ctrl_diff <- ctrl_post - ctrl_pre
tr_diff   <- tr_post - tr_pre
pre_diff  <- tr_pre - ctrl_pre
post_diff <- tr_post - ctrl_post

## Calculate the actual DiD Estimate
did_estimate <- tr_diff - ctrl_diff

## Assemble into a clean, labeled 3x3 matrix
did_table <- matrix(
  c(ctrl_pre,  ctrl_post,  ctrl_diff,
    tr_pre,    tr_post,    tr_diff,
    pre_diff,  post_diff,  did_estimate),
  nrow = 3, byrow = TRUE,
  dimnames = list(
    Group = c("Control Group (A)", "Treatment Group (B)", "Difference (B - A)"),
    Period = c("Pre-Period (time=0)", "Post-Period (time=1)", "Time Difference (Post-Pre)")
  )
)

## Print table with standard formatting (preventing scientific notation)
options(scipen = 999)
print(round(did_table, 2))

# Create Line Chart
## Calculate yearly averages for each group
chart_data <- df %>%
  group_by(year, treated) %>%
  summarize(mean_y = mean(y, na.rm = TRUE), .groups = 'drop') %>%
  mutate(Group = factor(treated, levels = c(0, 1), labels = c("Control Group", "Treatment Group")))

## Build the line chart
did_plot <- ggplot(chart_data, aes(x = year, y = mean_y, color = Group, group = Group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  
  ### Add vertical treatment intervention line between 1993 and 1994
  geom_vline(xintercept = 1994, linetype = "dashed", color = ftr_red, linewidth = 1) +
  
  ### Add descriptive label for the vertical line
  annotate("text", x = 1994.2, y = max(chart_data$mean_y)*1.1, 
           label = "Treatment Introduced", 
           color = ftr_red, fontface = "bold", hjust = 0) +
  
  ### Formatting labels and scales
  scale_x_continuous(breaks = min(chart_data$year):max(chart_data$year)) +
  scale_y_continuous(labels = scales::comma) + # Format y-axis numbers cleanly
  scale_color_manual(values = c("Control Group" = ftr_lt_blue, "Treatment Group" = ftr_md_blue)) +
  
  
  ### Labels and themes
  labs(
    title = "Difference-in-Differences Visual Trend",
    #subtitle = "...",
    #y = "..."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.title = element_blank(),
    axis.title = element_blank()
  )

  ### Render plot in RStudio/device window
  print(did_plot)

```

## SOURCES
 > https://www.princeton.edu/~otorres/DID101.pdf (DID overview)
 > https://www.publichealth.columbia.edu/research/population-health-methods/difference-difference-estimation (DID Overview)
 > https://www.princeton.edu/~otorres/DID101R.pdf (Example below)
 > https://bcallaway11.github.io/did/articles/did-basics.html (did package documentation)





